Iterative Monomorphisation

### 2025-06-25 iterative_monomorphisation.tex
shorten

Automatic theorem provers provide automation for proof assistant users. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One way to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}. However, this entails a lot of work that must be redone for every prover.

The alternative is to translate polymorphic problems to monomorphic problems.
One approach \cite[Section~9]{mono-trans} is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic. Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

@@

In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}). We also provide a more detailed\pagebreak[2] \hbox{description} to help future implementers. In addition, our paper shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).

Briefly, the algorithm works as follows. We assume problems to be sets of formulae. All the problem's symbols are collected, and the polymorphic symbol instances are matched against the monomorphic ones. This yields new symbol instances, both polymorphic and monomorphic. The process is then iterated, making use of the newly generated instances.

Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc., can be generated. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.
This example also shows that an infinite set of instances can be generated.

@@

We implemented iterative monomorphisation in Zipperposition~\cite{vukmirovic-et-al-2021}, a higher order prover written in OCaml. Although Zipperposition is polymorphic, it uses the monomorphic prover E~\cite{e} as a backend. Thanks to our work, E can now be used with polymorphic problems. Moreover, our implementation can be used as a preprocessor to interface with other stand-alone provers.
Our source code is available online.%
\footnote{\url{https://github.com/nartannt/iterative_monomorphisation}}

Our empirical evaluation on TPTP~\cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

@@

\item Is iterative monomorphisation more effective than the native polymorphism implemented in polymorphic provers?
\end{enumerate}
%
Our raw evaluation data is available online.%
\footnote{\url{https://zenodo.org/records/14881532}}

@@

Our algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any standard variant of rank-1 polymorphic logic. In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
% implemented by several automatic provers.

@@

This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
%Although the example may seem contrived, similar problems frequently arise in practice. For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.
A more elaborate example is provided in Appendix~\ref{app}.%
\footnote{The appendix example will be included here in the camera-ready version if
the program chairs give use an extra page or two, as is customary.}

@@

@@

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first order and higher order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems
that could be solved by the prover in at most 30 seconds
%(including monomorphisation when relevant)
per problem with a single thread.

@@

\subsection{Parameter optimisation}
\label{param_opti}

@@

\subsection{E as a Zipperposition backend}

@@

\subsection{Monomorphisation as a preprocessor}

@@

   We thank Sascha Böhme for fruitful discussions.
   We thank Jannis Limperg, Mark Summerfield, and the anonymous reviewers for suggesting textual improvements.

   This research is co-funded by the European Union (ERC, Nekoka, 101083038). Views and opinions expressed are however those of the authors only and do not necessarily reflect those of the European Union or the European Research Council. Neither the European Union nor the granting authority can be held responsible for them.
\end{credits}

@@

\appendix

\section{Detailed Example}
\label{app}

\begin{examplex}\rm
This example runs an example through the detailed version of the iterative algorithm works. Consider the following initial problem:
\begin{itemize}
   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
\end{itemize}
%
\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\). Then we enter the main loop (assuming \Loop is at least 1). We iterate over each formula and call \FStep for each of them. Nothing happens for formula \(\langle1\rangle\) because it contains no type variables. For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.

\SubstGen returns and the substitutions are used to generate new type argument tuples. Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
\begin{table}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
    \bottomrule
\end{tabular}
\end{table}
\vspace{-0.5em}

First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\). If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\). This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\). Now \(\sigma_2\) is selected, it instantiates both \(\alpha\) and \(\beta\) at once. In our case, \(\sigma_2\) is already in the set of monomorphising substitutions and is therefore ignored. \(\sigma_3\) is treated similarly. No more than \SubstLimit monomorphising substitutions can be generated this way.

Finally, all monomorphic formulae, both new and old, are mangled. This step is unecessary if the target language allows type constructors. In our example two new monomorphic formulae are generated. The algorithm outputs:
\vspace{-0.5em}
\begin{enumerate}
   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
\end{enumerate}

>>>>

Automatic theorem provers provide automation for proof assistant users. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank~1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers only work with monomorphic logics. One way to close this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}. However, this entails a lot of work that must be redone for every prover.
%
The alternative is to translate polymorphic problems to monomorphic problems.
One approach \cite{mono-trans} is to encode a polymorphic type system using dedicated function or predicate symbols in a monomorphic logic. Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}. This method heuristically instantiates the formulae's type variables with concrete types.

@@

In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}). We also provide a more detailed\pagebreak[2] \hbox{description} to help future implementers. In addition, we present some optimisations
to curb the combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
% some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).

The algorithm works as follows. We assume problems to be sets of formulae. All the problem's symbols are collected, and the polymorphic symbol instances are matched against the monomorphic ones. This yields new symbol instances, both polymorphic and monomorphic. The process is then iterated, making use of the newly generated instances.
%
Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc., can be generated. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.
This example also shows that an infinite set of instances can be generated.

@@

We implemented iterative monomorphisation in Zipperposition~\cite{vukmirovic-et-al-2021}, a higher order prover written in OCaml. Although Zipperposition is polymorphic, it uses the monomorphic prover E~\cite{e} as a backend. Thanks to our work, E can now be used with polymorphic problems. Moreover, our implementation can be used as a preprocessor for other stand-alone provers.
Our source code is available online.%
\footnote{\url{https://github.com/nartannt/iterative_monomorphisation}}
%
Our evaluation on the TPTP~\cite{tptp} tries to answer three questions (Section~\ref{sec:evaluation}):
\begin{enumerate}
\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?

@@

\item Is iterative monomorphisation more effective than the native polymorphism implemented in polymorphic provers?
\end{enumerate}

@@

Our algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any standard variant of rank~1 polymorphic logic. In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
% implemented by several automatic provers.

@@

This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
%Although the example may seem contrived, similar problems frequently arise in practice. For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.

@@

\section{Detailed example}

We run an example through the detailed version of the iterative algorithm. Consider this initial problem:
\begin{itemize}
   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
\end{itemize}
%
\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\). Then we enter the main loop (assuming \Loop is at least 1). We iterate over each formula and call \FStep for each of them. Nothing happens for formula \(\langle1\rangle\) because it contains no type variables. For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.

\SubstGen returns and the substitutions are used to generate new type argument tuples. Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
\begin{table}
\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
    \bottomrule
\end{tabular}
\end{table}

First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\). If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\). This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\). Now \(\sigma_2\) is selected, it instantiates both \(\alpha\) and \(\beta\) at once. In our case, \(\sigma_2\) is already in the set of monomorphising substitutions and is therefore ignored. \(\sigma_3\) is treated similarly. No more than \SubstLimit monomorphising substitutions can be generated in this way.

Finally, all monomorphic formulae, both new and old, are mangled. This step is unecessary if the target language allows type constructors. In our example two new monomorphic formulae are generated. The algorithm outputs:
\begin{enumerate}
   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
\end{enumerate}

@@

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in TF1 and TH1, corresponding respectively to first and higher order logic with rank~1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems
that could be solved by the prover in at most 30 seconds
%(including monomorphisation when relevant)
per problem with a single thread.
Our raw evaluation data is available online.%
\footnote{\url{https://zenodo.org/records/14881532}}

@@

\paragraph{\upshape\bfseries Parameter optimisation.}

@@

\paragraph{\upshape\bfseries E as a Zipperposition backend.}

@@

\paragraph{\upshape\bfseries Monomorphisation as a preprocessor.}

@@

   We thank Sascha Böhme for fruitful discussions.
   We thank Jannis Limperg, Mark Summerfield, and the peer reviewers for suggesting textual improvements.
   This research is co-funded by the European Union (ERC, Nekoka, 101083038). Views and opinions expressed are however those of the authors only and do not necessarily reflect those of the European Union or the European Research Council. Neither the European Union nor the granting authority can be held responsible for them.
\end{credits}

@@
